\chosentitle

### 2024-09-17 paper/main.tex
review abstract, intro, and methods

@@

@@

\begin{abstract}\noindent
Forward modelling stellar spectra usually requires a spectral synthesis code, or some non-linear interpolator constructed from a curated training set.
These approaches require pre-processing steps (e.g., continuum rectification) which, when performed separately, can bias subsequent measurements.
Here we present a \emph{linear} model that simultaneously fits stellar absorption, telluric transmission, and the joint continuum-instrument response.
Stellar absorption and telluric transmission is each modelled by factorizing a grid of rectified theoretical spectra into two non-negative matrices: basis weights and basis vectors.
The non-negativity constraint ensures that basis vectors are strictly additive.
The joint continuum-instrument response is modelled with a Fourier basis. 
Together this set of bases represents a very effective linear model for stellar spectroscopy, where the linearity ensures that inference is convex, stable, and fast.
This model allows us to reliably fit nuisance parameters (e.g., tellurics, continuum, radial velocity, rotational or macroscopic broadening), without any prior knowledge about the fundamental stellar properties.
We demonstrate our method by fitting \eso/\harps\ high-resolution echelle spectra of OBAFGKM-type stars.
We show from repeat observations of $\alpha$-Centauri A that our approach yields continuum estimates that are consistent to 0.2\% at S/N $\sim$ 100, and better than 0.5\% at S/N $\sim$ 30.
\end{abstract}

@@

Continuum normalization is often a necessary step before estimating stellar parameters and chemical abundances.
This process is ripe with subjectivity.
While there is general agreement in the literature that consistent continuum normalization is important, there is no apparent consensus on how it should be done.\\

Fitting the continuum correctly requires you to know where there is line absorption. But knowing where there is line absorption requires you to (at least) know the stellar parameters. Without knowing the stellar parameters -- or having a good model for line absorption -- spectroscopists have developed various bespoke methods. A popular choice is to iteratively mask pixels in an asymmetric manner (so-called `sigma clipping') to exclude data points some chosen level below the current estimate of the continuum, or to mask all pixels except a carefully selected set of so-called continuum pixels. However, the set of continuum pixels is only valid for stars of a similar spectral type and metallicity, and in many cases there are \emph{no} continuum pixels (e.g., M-dwarfs).\\

Classical spectroscopists might estimate some coarse continuum for the spectrum (or for each order, in the case of high-resolution echelle spectra), and then refine the continuum for every measured absorption line. This process is still often done by hand (albeit they are often experienced hands). Industrial spectroscopists might aim for a \emph{consistent} continuum normalization procedure: one that estimates a reliable continuum level for stars of similar stellar parameters and signal-to-noise ratios, even if that continuum estimate is some pseudo-continuum, and not the true continuum.\\

In an ideal scenario the continuum might be fit jointly with the stellar parameters, but this is often too expensive if it is costly to predict the emergent spectrum. In Section~\ref{sec:methods} we describe a family of methods to address these problems.
The methods we describe use non-negative matrix factorization (NMF) to approximate line absorption. NMF is a linear model to describe large non-negative matrix by two smaller matrices, both with non-negative elements. This non-negativity provides a very useful constraint that is applicable in many areas of astronomy (i.e., where things cannot physically be negative), but NMF has seen relatively little use in astronomy compared to other research areas, or other dimensionality reduction techniques.
We discuss limitations and potential extensions of our work in Section~\ref{sec:discussion}, before concluding in Section~\ref{sec:conclusions}.\\

@@

We will assume a forward model that includes three components: one that represents continuum-normalized stellar absorption (e.g., line absorption); a second to describe telluric transmission; and a third to represent the smooth continuum.\footnote{Throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response. These are different things that enter multiplicatively, but cannot be disentangled without extra work.} We will require all components to be linear models, which ensures that inference is stable and fast. In practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalization and handling other nuisances.\\

@@

Here we will describe the method in general before outlining explicit assumptions and implementation details. The data are a one-dimensional spectrum with $P$ pixels, where the $i$-th pixel has wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_{y_i}$ (with $1 \leq i \leq P$). The forward model for the flux in the $i$-th pixel can be expressed as the element-wise (Hadamard; $\hadamard$) multiplication of what we will describe as the stellar absorption model $f(\lambda_i; \vecalpha)$, the telluric transmission model $g(\lambda_i; \vecbeta)$, and the continuum-instrument response model $h(\lambda_i;\vecgamma)$
\begin{align}
    y_i &= f(\lambda_i;\vecalpha)\hadamard{}g(\lambda_i;\vecbeta)\hadamard{}h(\lambda_i;\vecgamma) + \mbox{noise}
\end{align}
where the components $f(...)$, $g(...)$, and $h(...)$ are defined below. Throughout this paper we fit in (natural) log-transformed data space $\vec{Y} = \log{\vec{y}}$ with the transformed variance in the $i$-th pixel
\begin{eqnarray}
    \vec{C}_{ii} = \left(\frac{\sigma_{y,i}}{y_i} - \frac{\sigma_{y,i}^2}{2y_i^2} + \frac{2\sigma_{y,i}^3}{8y_i^3} - \frac{6\sigma_{y,i}^4}{24y_i^4}\right)^2
\end{eqnarray}
\noindent{}computed by taking a swiftly Taylor series expansion. The log-transformation changes the element-wise product into a helpful summation:
\begin{align}
    \label{eq:log_y}
    \vec{Y} &= \log{f(\vec{\lambda}; \vecalpha)} + \log{g(\vec{\lambda};\vecbeta)} + \log{h(\vec{\lambda};\vecgamma)} \quad .
\end{align}\\

@@

We refer to our $N \times D$ matrix of continuum-rectified theoretical stellar spectra as $\vec{S}$. This is a dense matrix: there are no entries of exactly zero, with many entries near 1 (no absorption). However, a small transformation to this matrix can make it it extremely sparse. Numerous transformations are possible\footnote{Another sparse transformation is $1 - \vec{M}$, but this makes our resulting model non-linear.}, but for many reasons we chose to factorize the negative logarithm of $\vec{S}$ into two smaller matrices $\vec{W}_\star$ and $\vec{F}$ such that,
\begin{eqnarray}
    \label{eq:nmf}
    -\log\left({\vec{S}}\right) \approx \vec{W}_\star\transpose\vec{F}
\end{eqnarray}
where $\vec{W}_\star$\footnote{The star in $\vec{W}_\star$ differentiates it from the basis weights found from factorizing telluric spectra $\vec{W}_\earth$.} is a $K \times N$ matrix of \emph{basis weights}, where $K$ is the number of chosen basis vectors, and $\vec{F}$ is a $K \times D$ matrix of NMF \emph{basis vectors}. All elements in $-\log\left({\vec{S}}\right)$, $\vec{W}_\star$, and $\vec{F}$ are required to be non-negative. The number of basis components $K$ should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.\\%Figure~\ref{fig:schematic} illustrates some example basis vectors factorized from theoretical spectra in the \apogee\ wavelength range.\\

@@

The factorization of $-\log\left({\vec{S}}\right)$ into $\vec{W}_\star\transpose\vec{F}$ is reasonably fast and easy to compute given existing packages in modern programming languages. We found substantial improvements by initializing $\vec{F}$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $-\log\left({\vec{S}}\right)$: this is the default behaviour in many packages. We found that factorizing $\vec{W}_\star\transpose\vec{F}$ by coordinate descent was substantially faster than using multiplicative updates, although the multiplicative update rules are appealing because they are convex, do not require many choices, and they make for an efficient problem for graphical processing units.\\

There are very few requirements of the theoretical spectral grid. There are no requirements on the number of dimensions (e.g., whether or not to include $[\alpha/\mathrm{Fe}]$, $[\mathrm{C/Fe}]$), and no strict requirements (see Section~\ref{sec:discussion}) on spacing in between points. The only implicit requirement is that the theoretical spectra should approximately span the range of stars that you intend to fit. This is more of a recommendation than a requirement: in practice we found that a grid trained on theoretical spectra of OBA-type stars was also sufficiently flexible to model many (but not all) white dwarf spectra. For these reasons, we chose to include as many theoretical spectra as our memory constraints would allow.\footnote{If the number of spectra exceeds your memory constraints then there are numerous strategies available: you can use memory-mapped arrays, use lower precision float types, compute updates in batches, or simply skip every $n$th spectrum.}\\

We used the BOSZ spectral grid to construct basis vectors for a stellar line absorption model. This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid that spans a reasonably large range in stellar parameters, covers the complete wavelength range of the \eso/\harps\ instrument, and importantly it includes the theoretical continuum for each spectrum. We clipped any BOSZ normalized flux values exceeding 1 (e.g., emission lines) as they violate our NMF requirements. As we discuss in later sections, emission lines are sufficiently rare in stellar spectra that they do not contribute significantly to the $\chi^2$ at inference time. We used \todo{all BOSZ spectra and factorized with CD into K basis vectors for 1000 steps, at which point the reconstruction loss was X}.\\

\begin{figure*}
    \caption{A schematic showing some basis vectors computed by non-negative matrix factorization from a grid of theoretical stellar spectra. \label{fig:schematic}}
\end{figure*}

\todo{Discussion of the stellar and telluric basis vectors, how they separate out into astrophysically relevant things, and refer to the figure.}

@@

Telluric transmission can be included in a similar way to the stellar line absorption. We factorize the negative log of theoretical telluric absorption into two-non negative matrices $\vec{W}\transpose_\earth$ and $\vec{G}$. \todo{Details about what telluric grid we used, and setting used for factorizing. Note how many vectors we chose ($L$).} With the non-negative matrix $\vec{G}$, the function $\log{g\left(\lambda_i;\vecbeta\right)}$ is simply
\begin{eqnarray}
    \log{g(\lambda_i;\vecbeta)} = -\vec{G}\transpose\vecbeta
\end{eqnarray}
where $\vecbeta \in [0, \infty)$ is column vector of $L$ \emph{telluric basis weights} to be solved at inference time. We could restrict $\vecbeta \in [0, \max(\vec{W}_\earth)]$, and $\vecalpha \in [0, \max(\vec{W}_\star)]$, but in practice this is not necessary, and a more flexible approach might be to apply a feature weighting prior (see Section~\todo{?}). Like $\vecalpha$, $\vecbeta$ is analogous to a column in $\vec{W}_\earth$: it represents the $L$ basis weights needed to reconstruct the telluric absorption.\\

There are many suitable choices for the logarithm of the continuum-instrument response model $\log{h(\lambda_i;\vecgamma)}$. Here we chose a Fourier basis of sine and cosine functions because it is a linear representation, and is sufficiently flexible for modelling the joint continuum-instrument response across a variety of spectrographs. The component $\log{h(\lambda_i;\vecgamma)}$ is expressed compactly as
\begin{align}
    \log{h(\lambda_i;\vecgamma)} = \vec{H}\vecgamma
\end{align}
where $\vec{H}$ is a design matrix where the elements of the $j$th column are, % todo: shape of design matrix
\begin{align}
    \vec{H}_{j}(\lambda_i) & = \left\{\begin{array}{cl}\displaystyle\cos\left(\frac{\pi\,[j-1]}{P}\,\lambda_i\right) & \mbox{for $j$ odd} \\[3ex]
    \displaystyle\sin\left(\frac{\pi\,j}{P}\,\lambda_i\right) & \mbox{for $j$ even}\end{array}\right. ~,
\end{align}
\noindent{}where $P$ is a length scale which we set to be twice the length of the wavelength range to fit (e.g., twice the peak-to-peak range of an echelle order). The design matrix $\vec{H}$ can be constructed \emph{a priori} before inference begins. Throughout this paper we will describe $\vecgamma$ as the sine-and-cosine \emph{coefficients}.\\

\noindent{}With $f(\lambda_i;\vecalpha)$, $g(\lambda_i;\vecbeta)$ and $h(\lambda_i;\vecgamma)$ now defined, we can expand the forward model for the log-transformed data $\vec{Y}$,
\begin{equation}
    \vec{Y} = -\vec{F}\transpose\vecalpha - \vec{G}\transpose\vecbeta + \vec{H}\vecgamma
\end{equation}
\begin{equation}
    \vec{Y} = \vec{A}\vec{X}
\end{equation}
where the parameters and design matrices are stacked to construct $\vec{A}$ and $\vec{X}$:
\begin{eqnarray}
    \vec{A} = \begin{bmatrix}-\vec{F}\transpose\\-\vec{G}\transpose\\\vec{H}\end{bmatrix}
    \quad \mbox{and} \quad
    \vec{X} = \begin{bmatrix}\vecalpha\\\vecbeta\\\vecgamma\end{bmatrix} \quad .
\end{eqnarray}

Given some log-transformed observed flux $\vec{Y}$ and associated covariance matrix $\vec{C}$, this reduces to a weighted least-squares problem. This could be solved directly by linear algebra,
\begin{eqnarray}
    \vec{\tilde{X}} = (\vec{A}\transpose\vec{C}^{-1}\vec{A})^{-1}\vec{A}\transpose\vec{C}^{-1}\vec{Y}
\end{eqnarray}
\noindent{}but this does not guarantee that the bounds $\vecalpha \geq 0$ and $\vecbeta \geq 0$ are met. There are a few ways to find a solution to this system that is within the parameter bounds (e.g., linear operators; expectation-maximization), and thankfully all of them are convex paths. For simplicity we use the truncated reflective trust region algorithm, where we provide the Jacobian ($\vec{A}$).\\

@@

After solving for $\vec{\tilde{X}}$, the (un-transformed) rectified flux and continuum can be computed in a straightorward manner:
\begin{eqnarray}
    \mathrm{rectified~stellar~flux}, \quad &f& = \exp{\left(-\vec{F}\transpose\boldsymbol{\tilde{\alpha}}\right)} \\
    \mathrm{telluric~transmission},\quad  &g& = \exp{\left(-\vec{G}\transpose\boldsymbol{\tilde{\beta}}\right)} \\
    \mathrm{continuum}, \quad &h& = \exp{\left(\vec{H}\boldsymbol{\tilde{\gamma}}\right)}
\end{eqnarray}

>>>>

\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

@@

% TODO
% [ ] Talk about BOSZ spectra and how the model was constructed.
% [ ] Talk about how we constructed the telluric basis vectors
% [ ] Flesh out discussion points
% [ ] Data section
% -> Blaze???
% [ ] Results section
% -> ON a set of OBAFGKM stars, provides clean and consistent normalisation across these types
% -> ARE WE USING THE NIGHTLY BLAZE?
% -> discuss for alf Cen A, how we selected spectra to get a uniform sampling in SNR, and how we fit them
% -> how we define the consistency measure

% [ ] discussion
% -> We fix vsini / v_rel from the literature, but these can be computed 
% ->

@@

\begin{abstract}\noindent
Forward modelling stellar spectra usually requires a spectral synthesis code, or some non-linear interpolator constructed from a curated training set.
These approaches require pre-processing steps (e.g., continuum rectification) which, when performed separately, can bias subsequent measurements.
Here we present a \emph{linear} model that simultaneously fits stellar absorption, telluric transmission, and the joint continuum-instrument response.
Stellar absorption and telluric transmission are each modelled by factorizing a grid of rectified theoretical spectra into two non-negative matrices: basis weights and basis vectors.
The non-negativity constraint ensures that basis vectors are strictly additive.
The joint continuum-instrument response is modelled with a Fourier basis. 
Together this set of bases represents a very effective linear model for stellar spectroscopy, where the linearity ensures that inference is convex, stable, and fast.
This model allows us to reliably fit nuisance parameters (e.g., tellurics, continuum, radial velocity, rotational or macroscopic broadening), without any prior knowledge about the fundamental stellar properties.
We demonstrate our method by fitting \todo{toy models of M-dwarf spectra where we recover the correct continuum, and by fitting} \eso/\harps\ high-resolution echelle spectra of OBAFGKM-type stars.
From repeat observations of $\alpha$-Centauri A we show that our approach yields continuum estimates that are consistent to 0.2\% at S/N $\sim$ 100, and better than 0.5\% at S/N $\sim$ 30.
\end{abstract}

@@

Continuum normalization\footnote{Or continuum rectification, as it was historically described.} is often required before estimating stellar parameters and chemical abundances, because these quantities are inferred from line strengths measured \emph{relative} to the continuum.
The practicalities of continuum rectification is ripe with subjectivity, for justified historical reasons: often a consistent procedure is more useful than an abstract one, so people tend to stick with what they know. While there is general agreement in the literature that consistent continuum normalization is important, there is no apparent consensus on how it should be done.\\

There is a variety of rectification strategies used to achieve different goals. Classical spectroscopists might make some coarse estimate of the continuum for the spectrum, and locally refine the continuum for each sbsorption line of interest. This process is often done by hand,\footnote{Albeit they are often experienced hands. See \citet{Bensby:2014}: \quote{`\emph{more than 300\,000 equivalent widths were measured by (the first author's right) hand}'}.} although some graphical user interface tools exist to streamline the process and improve reproducibility. This approach usually restricts the spectroscopist to only measuring strengths of isolated lines: those that are not blended by neighbouring atomic lines or strong molecular bands. In contrast, industrial spectroscopists have a different objective: one where they solely for a \emph{consistent} continuum normalization procedure. For them the process does not have to deliver the true continuum, but it should deliver a \emph{pseudo}-continuum that is consistent for stars of similar stellar parameters and signal-to-noise (S/N) ratios.\\

Fitting the continuum correctly requires you to know where there is absorption. But knowing where there is line absorption requires you to (at least) know the stellar parameters. Without knowing the stellar parameters -- or having a good model for line absorption -- spectroscopists have developed various bespoke methods. A popular choice is to iteratively mask pixels in an asymmetric manner (so-called `sigma clipping') to exclude data points some chosen level below the current estimate of the continuum. This works well if the practictioner can confidently set the bounds on where to exclude data. However, those parameters would need to be different for a more metal-rich star, or for a spectrum with a low signal-to-noise (S/N) ratio. Exactly how to specify the asymmetric clipping parameters comes down to experience.\\

An alternative is simply to mask all pixels except a carefully selected set of so-called `continuum pixels'. This is painstaking work to perform manually, but can be done by iteratively training data-driven models, or building masks from approximate line strengths computed from atomic properties. In either scenario, the set of continuum pixels is only valid for stars of a similar type and metallicity. In many cases, there are simply \emph{no} continuum pixels (e.g., near a molecular absorption band in an M-dwarf), and the best we can compute is a \emph{pseudo}-continuum.\\

A variety of methods for rectifying spectra are described in the literature. \todo{Some summary of those. Includes the rolling ball thing, RASSINE, and any others?}\\

In an ideal scenario the continuum is jointly fit with the stellar parameters. Unfortunately, predicting the emergent spectrum is usually expensive. In Section~\ref{sec:methods} we describe a family of methods to address these problems.
The methods we describe use non-negative matrix factorization (NMF) to approximate line absorption. NMF is a linear model to describe large non-negative matrix by two smaller matrices, both with non-negative elements. This non-negativity provides a very useful constraint that is applicable in many areas of astronomy (i.e., where things cannot physically be negative), but NMF has seen relatively little use in astronomy compared to other research areas, or other dimensionality reduction techniques.
We discuss limitations and potential extensions of our work in Section~\ref{sec:discussion}, before concluding in Section~\ref{sec:conclusions}.\\

@@

We will assume a forward model that includes three components: one that represents continuum-normalized stellar absorption (e.g., molecular or atomic line absorption); a second to describe telluric transmission; and a third to represent the smooth continuum.\footnote{Throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response. These are different things that enter multiplicatively, but cannot be disentangled without extra work.} We will require all components to be linear models, which ensures that inference is stable and fast. In practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalization and handling other nuisances.\\

@@

Here we will describe the method in general before outlining the implementation details. We assume the data are a one-dimensional spectrum with $P$ pixels, where the $i$-th pixel has wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_{y_i}$ (with $1 \leq i \leq P$). The forward model for the flux in the $i$-th pixel can be expressed as the element-wise (Hadamard; $\hadamard$) multiplication of what we will describe as the stellar absorption model $f(\lambda_i; \vecalpha)$, the telluric transmission model $g(\lambda_i; \vecbeta)$, and the continuum-instrument response model $h(\lambda_i;\vecgamma)$
\begin{align}
    y_i &= f(\lambda_i;\vecalpha)\hadamard{}g(\lambda_i;\vecbeta)\hadamard{}h(\lambda_i;\vecgamma) + \mbox{noise}
\end{align}
where the components $f(...)$, $g(...)$, and $h(...)$ are defined below. Throughout this paper we fit in (natural) log-transformed data space $\vec{Y} = \log{\vec{y}}$ with the transformed variance in the $i$-th pixel
\begin{eqnarray}
    \vec{C}_{ii} = \left(\frac{\sigma_{y,i}}{y_i} - \frac{\sigma_{y,i}^2}{2y_i^2} + \frac{2\sigma_{y,i}^3}{8y_i^3} - \frac{6\sigma_{y,i}^4}{24y_i^4}\right)^2
\end{eqnarray}
\noindent{}computed by taking a swiftly Taylor series expansion. The log-transformation changes the element-wise product into a convenient summation:
\begin{align}
    \label{eq:log_y}
    \vec{Y} &= \log{f(\vec{\lambda}; \vecalpha)} + \log{g(\vec{\lambda};\vecbeta)} + \log{h(\vec{\lambda};\vecgamma)} \quad .
\end{align}\\

@@

We refer to our $N \times D$ matrix of continuum-rectified theoretical stellar spectra as $\vec{S}$. This is a dense matrix: there are no entries of exactly zero, with many entries near 1 (no absorption). However, a small transformation to this matrix makes it extremely sparse. Numerous transformations are possible\footnote{Another sparse transformation is $1 - \vec{M}$, but this makes our resulting model non-linear.}, but for many reasons we chose to factorize the negative logarithm of $\vec{S}$ into two smaller matrices $\vec{W}_\star$ and $\vec{F}$ such that,
\begin{eqnarray}
    \label{eq:nmf}
    -\log\left({\vec{S}}\right) \approx \vec{W}_\star\transpose\vec{F}
\end{eqnarray}
where $\vec{W}_\star$\footnote{The star in $\vec{W}_\star$ differentiates it from the basis weights found from factorizing telluric spectra $\vec{W}_\earth$.} is a $K \times N$ matrix of \emph{basis weights}, where $K$ is the number of chosen basis vectors, and $\vec{F}$ is a $K \times D$ matrix of NMF \emph{basis vectors}. The factorization of $-\log{\vec{S}}$ into $\vec{W}_\star\transpose\vec{F}$ requires that all elements in $-\log\left({\vec{S}}\right)$, $\vec{W}_\star$, and $\vec{F}$ be non-negative. The number of basis components $K$ should be significantly smaller than both the number of input spectra $N$ and the number of pixels $D$ per spectrum.\\%Figure~\ref{fig:schematic} illustrates some example basis vectors factorized from theoretical spectra in the \apogee\ wavelength range.\\

@@

The factorization of $-\log\left({\vec{S}}\right)$ into $\vec{W}_\star\transpose\vec{F}$ is reasonably fast and easy to compute given existing packages in modern programming languages. In general we found substantial improvements by initializing $\vec{F}$ with non-negative double singular value decomposition, where zeros were filled-in with the average of $-\log\left({\vec{S}}\right)$: this is the default behaviour in many packages. We found that factorizing $\vec{W}_\star\transpose\vec{F}$ by coordinate descent was substantially faster than using multiplicative updates, \todo{until reaching a plateau in reconstruction error, at which point we would iterate further using multiplicative updates.} The multiplicative update rules are appealing because they are convex, do not require many choices, and they can be efficiently iterated on a graphics card.\\

There are very few requirements of the theoretical spectral grid. There are no requirements on the number of dimensions (e.g., whether or not to include $[\alpha/\mathrm{Fe}]$, $[\mathrm{C/Fe}]$), and no strict requirements (see Section~\ref{sec:discussion}) on spacing in between points. The only implicit requirement is that the theoretical spectra should approximately span the range of stars that you intend to fit. This is more of a recommendation than a requirement: in practice we found that a grid trained on theoretical spectra of A- to M-type stars was also sufficiently flexible to model many O-type stars. For these reasons, we chose to include as many theoretical spectra as our memory constraints would allow.\footnote{If the number of spectra exceeds your memory constraints then there are numerous strategies available: you can use memory-mapped arrays, use lower precision float types, compute updates in batches, or simply skip every $n$th spectrum.}\\

We used the BOSZ spectral grid to construct basis vectors for a stellar line absorption model. This is a high-resolution ($\mathcal{R} \sim 300{,}000$) finely sampled theoretical grid computed using the \project{ATLAS9} code with \project{MARCS} models atmospheres. The grid spans a reasonably large range in stellar parameters, covers the complete wavelength range of the \eso/\harps\ instrument, and importantly it includes the theoretical continuum for each spectrum. We constructed a sparse kernel to convolve the rectified flux to the \eso/\harps\ spectral resolution of $\mathcal{R} \sim 115{,}000$, truncated the convolved spectra between 300-700\,nm, and clipped any rectified flux values exceeding 1 (e.g., emission lines) as they violate our NMF requirements. The matrix $\vec{S}$ includes all \todo{X} convolved BOSZ spectra, with \todo{N} pixels each, which we randomly shuffled before factorization. We then factorized this matrix into $K = 16$ basis vectors by coordinate descent for \todo{10,000} steps, \todo{and then X steps using multiplicative updates}. After training, the Frobenius loss, also known as the L2 reconstruction loss
\begin{eqnarray}
    L_2 = \norm{{\left(-\log{\vec{S}}\right) - \vec{W}_\star\transpose\vec{F}}}^2
\end{eqnarray}
\noindent{}was found to be \todo{X}. Figure~\ref{fig:schematic} shows some example basis vectors computed by this factorization, where it is somewhat striking that the basis vectors are identifiable by astrophysical mechanism, since no stellar parameters (or astrophysics of any kind) are included in the factorization process. For example, one basis vector just includes hydrogen lines. Another basis vector only has CO-molecule lines. We expand on this observation in later sections.\\

\begin{figure*}
    \includegraphics*[width=\textwidth]{nmf_plot.pdf}
    \caption{A schematic showing some basis vectors computed by non-negative matrix factorization from a grid of theoretical stellar spectra. \todo{Update this for BOSZ spectra, not APOGEE.}\label{fig:schematic}}
\end{figure*}

@@

Telluric transmission can be included in a similar way to stellar absorption. We used a grid of model telluric transmission spectra computed for the \texttt{PypeIt} project. These spectra are computed using the Line-by-Line Radiative transfer Model for an array of observatory locations (latitude, longitude) and atmospheric temperatures, pressures, humidity, and airmass. The original grid of telluric transmission spectra was not available to us. Instead, we used a representation of this grid that is shipped with \texttt{PypeIt}, and compressed using Principal Component Analysis and an inverse hyperbolic sine transformation, to approximate the original grid. With this reconstructed grid, we factorized the negative log of theoretical telluric absorption 
\begin{eqnarray}
    -\log\vec{M} \approx \vec{W}\transpose_\earth\vec{G}
\end{eqnarray}
\noindent{}into two-non negative matrices $\vec{W}\transpose_\earth$ and $\vec{G}$ using $L = 4$ basis vectors. We took \todo{10,000} iterations using coordinate descent, at which point the reconstruction loss was \todo{X}. With the non-negative matrix $\vec{G}$, the function $\log{g\left(\lambda_i;\vecbeta\right)}$ is simply
\begin{eqnarray}
    \log{g(\lambda_i;\vecbeta)} = -\vec{G}\transpose\vecbeta
\end{eqnarray}
where $\vecbeta \in [0, \infty)$ is column vector of $L$ \emph{telluric basis weights} in to be solved at inference time. 
%We could restrict $\vecbeta \in [0, \max(\vec{W}_\earth)]$, and $\vecalpha \in [0, \max(\vec{W}_\star)]$, but in practice this is not necessary, and a more flexible approach might be to apply a feature weighting prior (see Section~\todo{?}). 
Like $\vecalpha$, $\vecbeta$ is analogous to a column in $\vec{W}_\earth$: it represents the $L$ basis weights needed to reconstruct the telluric absorption.\\

There are many suitable choices for the logarithm of the continuum-instrument response model $\log{h(\lambda_i;\vecgamma)}$. Here we chose a Fourier basis of sine and cosine functions because it is a linear representation, and is sufficiently flexible for modelling the joint continuum-instrument response across a variety of spectrographs. The component $\log{h(\lambda_i;\vecgamma)}$ is expressed compactly as
\begin{align}
    \log{h(\lambda_i;\vecgamma)} = \vec{H}\vecgamma
\end{align}
where $\vec{H}$ is a design matrix where the elements of the $j$th column are, % todo: shape of design matrix
\begin{align}
    \vec{H}_{j}(\lambda_i) & = \left\{\begin{array}{cl}\displaystyle\cos\left(\frac{\pi\,[j-1]}{P}\,\lambda_i\right) & \mbox{for $j$ odd} \\[3ex]
    \displaystyle\sin\left(\frac{\pi\,j}{P}\,\lambda_i\right) & \mbox{for $j$ even}\end{array}\right. ~,
\end{align}
\noindent{}where $P$ is a length scale which we set to be twice the length of the wavelength range to fit (e.g., twice the peak-to-peak range of an echelle order). %The design matrix $\vec{H}$ can be constructed \emph{a priori} ibce before inference begins. 
Throughout this paper we will describe $\vecgamma$ as the sine-and-cosine \emph{coefficients}.\\

\noindent{}With $f(\lambda_i;\vecalpha)$, $g(\lambda_i;\vecbeta)$ and $h(\lambda_i;\vecgamma)$ now defined, we can expand the forward model for the log-transformed data $\vec{Y}$,
\begin{equation}
    \vec{Y} = -\vec{F}\transpose\vecalpha - \vec{G}\transpose\vecbeta + \vec{H}\vecgamma
\end{equation}
\noindent{}which we can compactly write as
\begin{equation}
    \vec{Y} = \vec{A}\vec{X}
\end{equation}
where the parameters and design matrices are stacked to construct $\vec{A}$ and $\vec{X}$:
\begin{eqnarray}
    \vec{A} = \begin{bmatrix}-\vec{F}\transpose\\-\vec{G}\transpose\\\vec{H}\end{bmatrix}
    \quad \mbox{and} \quad
    \vec{X} = \begin{bmatrix}\vecalpha\\\vecbeta\\\vecgamma\end{bmatrix} \quad .
\end{eqnarray}

Given some log-transformed observed flux $\vec{Y}$ and associated covariance matrix $\vec{C}$, this reduces to a weighted least-squares problem. This could be solved directly by linear algebra,
\begin{eqnarray}
    \vec{\tilde{X}} = (\vec{A}\transpose\vec{C}^{-1}\vec{A})^{-1}\vec{A}\transpose\vec{C}^{-1}\vec{Y}
\end{eqnarray}
\noindent{}but this solution is not subject to the constraints $\vecalpha \geq 0$ and $\vecbeta \geq 0$. There are a few ways to solve this system subject to the boundary constraints, but the choice is not very important because the problem is convex. For simplicity we use the truncated reflective trust region algorithm where we stop optimizing once the relative tolerance has improved by less than $10^{-16}$.\\

@@

After solving for $\vec{\tilde{X}}$, the (un-transformed) quantities can be easily computed:
\begin{eqnarray}
    \mathrm{rectified~stellar~flux}, \quad &f& = \exp{\left(-\vec{F}\transpose\boldsymbol{\tilde{\alpha}}\right)} \\
    \mathrm{telluric~transmission},\quad  &g& = \exp{\left(-\vec{G}\transpose\boldsymbol{\tilde{\beta}}\right)} \\
    \mathrm{continuum}, \quad &h& = \exp{\left(\vec{H}\boldsymbol{\tilde{\gamma}}\right)}
\end{eqnarray}
